---
title: "Tidy data"
subtitle: "Lesson 4: Manipulating data"
date: "September 2020"
output: 
  html_document:
    theme: flatly
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: true
---

## Goals of this lesson:
- View data with histograms 
- Exclude data with **filter()**
- Create new variables with **mutate()**
- Change variable names with **rename()**
- Subset variables with **select()**


## 0. Open packages 
```{r, warning = F, message = F}
# load packages 
library(readxl)
library(tidyverse)
```

## 1. load cortisol data 
```{r, message = F, warning = F}
cort_data <- read_csv("../data/cort_data_wide.csv")
```

## 2. Inspect data 

First, look at the data to s ee what we're working with.

We have four cortisol measurements. These represent waking and bedtime values collected across 2 days. The dates of cortisol collection are "cdate1" and "cdate2". 

We also have participant id and a categorical variable, caregiving type.
```{r}
glimpse(cort_data)
```

We also have a lot of missing data!
```{r}
sum(is.na(cort_data$cort_waking1))
sum(is.na(cort_data$cort_waking2))
```

### Visualize data with histogram

let's do a simple histogram:

```{r}
hist(cort_data$cort_bedtime2)
```

Two observations:  
1) There are extreme outliers in the cort data  
2) The data is not normally distributed 


## 3. Clean / filter the data

The standard cut-off for biologically plausible cortisol values is 2, so let's remove any value > 2. 

We previously used **filter()** to remove NAs, but it can work with any logical expression. 

**Note**: If you use conditional logic based on a value, and some rows have NA for that value, the NA rows will also be excluded by default! 
```{r}
cort_data_clean <- cort_data %>%
  filter(cort_waking1 < 2 & cort_waking1 > 0 &
           cort_waking2 < 2 & cort_waking2 > 0 &
           cort_bedtime1 < 2 & cort_bedtime1 > 0 &
           cort_bedtime2 < 2 & cort_bedtime2 > 0)
```


### How many participants were excluded? 

Let's make a little table by counting the rows of the raw data and cleaned data
```{r}
N_total <- nrow(cort_data)
N_included <- nrow(cort_data_clean)
N_excluded <- N_total - N_included

# data.frame() will combine the items into a dataframe
# which is just a little table here.
data.frame(N_total, N_included, N_excluded)
```

We had a lot of outliers! This is fake data, so we'll be ok with this. 

## 4. Create new variables

Let's log-transform the cortisol values to normalize the distribution. 

The **mutate()** function applies an operation to the input value, and generates a new column. 

```{r}
cort_data_clean <- cort_data_clean %>%
  # new variable  on left, transformation on the right
  mutate(log_cort_waking1 = log(cort_waking1), 
         log_cort_waking2 = log(cort_waking2), 
         log_cort_bedtime1 = log(cort_bedtime1),
         log_cort_bedtime2 = log(cort_bedtime2))
```

Let's check the distribution now.
```{r}
hist(cort_data_clean$log_cort_bedtime2)
```


## 5. Renaming variables 

While we're cleaning up the data, let's also rename some less-obvious variable names. 

```{r}
cort_data_clean <- cort_data_clean %>%
  # new name on left, old name is  on the  right
  rename(cort_date1 = cdate1, 
         cort_date2 = cdate2)
```

## 6. Selecting variables

Let's take a look at the data now.
```{r}
head(cort_data_clean)
```

We still have the raw cort values, and the logged values.
Let's clean up and remove the variables with raw values from the dataset. 

The **select()** function let's you subset variables that you want to maintain, or discard. Any variable names inside the parentheses will be kept, unless they have a minus sign, in which case they will be removed.  

Here, we're going to remove the raw cort variables (and keep everything else)

```{r}
cort_data_clean1 <- cort_data_clean %>%
  select(-cort_waking1, -cort_waking2, -cort_bedtime1, -cort_bedtime2)

# another way 
cort_data_clean0 <- cort_data_clean[, 6:ncol(cort_data_clean)]
```

###  A short-cut:

Selecting variables one-by-one can get very tedious when you have many columns.

The **starts_with()** function captures strings that start with the specific pattern you provide. It's automatically going to capture the 4 variables that start with "cort" (and remove them)
```{r}
cort_data_clean2 <- cort_data_clean %>%
  select(-starts_with("cort"))

names(cort_data_clean2)
```


## Save your data

First, make a new folder called "clean_data" in the Files panel.

```{r}
# this is the cort in wide format, with the demo data added.
write.csv(cort_data_clean1, "../clean_data/cort_clean_wide.csv", row.names = F)
```

