---
title: "Getting started with R"
subtitle: "Lesson 1: The 3 Rs (R studio, R Markdown & R Projects)"
date: "September 2020"
output: 
  html_document:
    theme: flatly
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: true
---

# Goals of this lesson:
- Install packages
- Introduction to R studio environment
- Introduction to R projects
- Introduction to R markdown 


# 1.  Brief tour of R Studio 

**Scripts:** These are individual files where you write code that you want to save.

**Console:** Code that's run gets executed in the console. 

**Environment:** Any objects, or data, that are loaded into the R session, or created during the R session, will appear here!

Run this chunk of code, and you'll see the object "two" appear in the environment, which consists of the value "2". 
```{r}
# This is a code chunk that excutes code 
two <- 1 + 1 
```

**Files:** Where you are on your computer, like finder within R studio.  If you're not already there, navigate to the directory where this file lives (click the three dots on the right hand side to open the file browser)

**Packages:** A bunch of handy functions bundled up together. 
Let's install some packages that we'll use in the tutorial.

```{r}
#install.packages(pkgs = c("readxl", "tidyverse", "knitr", "here"))
```

Packages are only installed once, but in order to access them, we need to load them into the current R session. 

It's good practice to load packages that you will need at the beginning of each script.

```{r}
library(readxl)
library(tidyverse)
library(knitr)
library(here)
```

# 2. R markdown 
R markdown is a flexible document where you can write formatted text, run code, and display the results. This is an R markdown document! 

**Why is R markdown useful for beginner R users?**

1. It gets you in the habit of organizing your code into discrete chunks.
2. It also encourages you to add comments to your code as you go.
3. In order to successfully "knit" the file, your code must be free of errors! So it helps you learn how to de-bug your code. 

**Code chunks:**  Where you write code to be executed. 
You can add a new code chunk by clicking +insert 

```{r}
# This is a code chunk that excutes code 
code = "my_code" 
```

**Markdown:** Style of formatting text, which you're seeing here. 
Any text written in the white space of an Rmd file is markdown.

**Knitting:** Generate a report from an R markdown file. 
We can turn this file into an html report or word document with a press of a button! 

1. hash out line 30 (installing packages)
2. click "knit" 
3. wait for the magic to happen!

The "knitted" html is automatically saved as the same name as this file, in the same folder, with the .html extension.

This is just the start! R markdown is highly customizable and you can generate html reports, manuscript drafts, or slides for a presentation. 

# 3. R projects 
Before explaining R projects, let's learn about working directories and file paths.

**Working directory:** The path to the current location where the R session is working from.  
```{r}
here()
```

Are you somewhere random on your computer? Your current working directory is probably where ever you saved this Rmd file. 

We need to know the working directory to open or save files.
For example, let's save our "two" object to a csv file.
By only including a file name, and no path, the file is automatically saved to the working directory. 
```{r}
write.csv(two, file = "two.csv")
```

What if we want to save this file to a specific folder? We can write out the **absolute path** to that exact folder, which can be long & prone to typos:

```{r,  eval = F}
# note: this won't work on your computer because the path doesn't exist!
write.csv(two, file = "~/Documents/Columbia/grad_school/research/masters_proj/cortisol/cort_part1/data/raw_data/two.csv")

```

Or, we can write the file path **relative** to the current working directory.  Run this chunk of code, and then navigate two folders back in the R studio Files window. 

```{r}
# let's save the file to folders back
write.csv(two, file = "../../two.csv")

```

Can you find the file on your computer?

Relative paths are also good for reproducibility... 

- If you need to run the code on a different machine 
- if you want to share your code with others, so they can run it on a different machine.

The **R project** is a placeholder that links the files in a specific directory.

**Make a new project:**

- File --> new project --> new directory --> new project
- Name your project and save it on your computer in a directory you'll remember. 

**Then, set up your project**

- File --> open file --> navigate to where you saved this lesson, and open it. 
- In the files window, make some new folders:
-- scripts/
-- data/
-- tables/
- save this file into the scripts folder!


# Additional Resources:
-  For more on R markdown, to customize your knitted files:
https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf
