---
title: "Getting started with R"
subtitle: "Lesson 2: Dataframe basics in R"
date: "September 2020"
output: 
  html_document:
    theme: flatly
    highlight: haddock 
    df_print: kable
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: true
---

## Goals of this lesson: 
- Loading data in R 
- Inspecting data 
- How to index dataframes
- Working with missing values


```{r, message = F, warning = F}

library(readxl)
library(tidyverse)
library(lubridate)
```

## 1. Loading data in R 

### Opening data 

You can open many different file types in R, as long as you use the right function to open the file. For example, **read_excel()** can be used to open excel files. 

We will open a csv file with **read_csv()**.

- Save the demo_data.csv into the "data" folder in your R project.

- The file path is relative to where this script is saved, so we need to go back up one level to the home folder, then into  the data folder: **"../data/demo_data.csv"**

- Then, we assign the file contents as **demo_data**. Now you have a dataframe object saved in the R environment.

```{r, warning = F , message = F}
# Note: File paths are always represented as character strings in quotes
demo_data <- read_csv("../data/demo_data.csv")

```


## 2. Inspect the data

### how many rows are in the data? 
```{r}
nrow(demo_data)
```

### how many columns are in the data? 
```{r}
ncol(demo_data)
```

###  print out the column names for each variable 
```{r}
names(demo_data)
```

### take a look at the first six rows of data

```{r}
head(demo_data)
```

### Variable types 
- Our data has 2 numeric variables
- and 3 categorical variables (characters). 
- We can also see that there are missing values (NA's)
```{r}
summary(demo_data)
```


### view entire dataframe in a spreadsheet 
OR you can navigate to the **Environment** pane and click on **demo_data** 
```{r}
#View(demo_data)

```


## 2. Indexing data
The data is stored as rows and columns, like a matrix. 
We can index the data with row and column indices:
**data[rows, columns]**

### show the first row of data
```{r}
demo_data[1,]
```

### show the first few rows of data
```{r}
demo_data[1:3,]
```

### show the first column of data, Participant_id
```{r}
demo_data[,1]
```


## 3. The power of the dataframe

The data is actually a special R object, called a "data.frame"
```{r}
class(demo_data)[4]
```

Instead of indexing with numbers, with data frames, we can index with column names.

This will make our life much easier when working with larger datasets and doing analysis. 

Let's start with a simple example to illustrate. How many participants are there in this dataset?

### show the Participant_id column
```{r}
demo_data$Participant_id
```

### Find the number of  participant IDs 
**unique()** is a useful function that returns only the unique items (e.g. removes any duplicates).
```{r}
unique(demo_data$Participant_id)
```

### Assign it as a new object 
The assign operator in R is <- 
```{r}
subject_list <- unique(demo_data$Participant_id)
```

### Find the length of that object
```{r}
length(subject_list)
```

### Find the number of rows in the data
```{r}
nrow(demo_data)
```

We can check if they're identical, by using **identical()** which returns TRUE or FALSE
```{r}
identical(length(subject_list), nrow(demo_data))

```

In this case, the number of subjects is the same as the number of rows, but this is a useful check when working with data that could have duplicates, or longitudinal data. 

## 4. Missing values 

In R, missing values are represented as **NA** - not 999 (MPlus) - or blanks (SPSS)

### Are there NAs in the age variable? 

The function **is.na()** returns a **logical vector**, where TRUE = is NA, and FALSE = not NA.
```{r}
is.na(demo_data$age)
```

### How many NAs are there in the age column? 
We can add logical vectors with **sum()**, as if they were a column of 1 and 0s, where True = 1 and False = 0.
```{r}
sum(is.na(demo_data$age))
```

### Which row has the NA in the age column?
We can use the logical vector to find the row index (number) where age = NA, and save it as an index. 
```{r}
NA_index <- which(is.na(demo_data$age))
```

### Which subject has NA in the age column? 
We can use that index to obtain the participant with missing age data. 
```{r}
demo_data$Participant_id[NA_index]
```

### Replace NA with a new value
We can use the assign operator to replace the NA with a value.
```{r}
# let's say that the NA  value should be  replaced with 52.
demo_data$age[NA_index] <- 52
```