---
title: "Tidy data"
subtitle: "Lesson 5: Merge and reshape data"
date: "September 2020"
output: 
  html_document:
    theme: flatly
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: true
---

## Goals of this lesson:
- Merge datasets together with **join()**
- Reshape data from wide to long with **pivot_longer()** 
- Reshape data from long to wide withs **pivot_wider()**
- String manipulation in character variables

## 0. load packages 
```{r, warning = F, message = F}
# load packages 
library(readxl)
library(tidyverse)
```

## 1. load data 

Open two datasets, the demographic data and the cleaned cortisol data.

```{r}
cort_data_clean <- read.csv ("../clean_data/cort_clean_wide.csv")

demo_data <- read.csv("../data/demo_data.csv")
```


## 2. Merge datasets 

Merging data, also called "joining" data, matches two datasets by "keys".

keys are variables that are in both datasets.

In this case, the key is 'participant_id'.

### Check your 'key' variable
Let's take a look at participant ID in each dataset: 
```{r}
head(demo_data$Participant_id)
```

This one isn't in the same format!
```{r}
head(cort_data_clean$Participant_id)
```

We are missing the "P" infront of the ID number.

Let's reformat Participant_id using **paste0()** and **mutate()**

Paste0() is a handy function pastes together strings!
```{r}
cort_data_clean <- cort_data_clean %>%
  mutate(Participant_id = paste0("P", Participant_id))

head(cort_data_clean$Participant_id)
```

### Perform the join / merge

**left_join()** will join two datsets, where the data on the left is the "base", and the data on the right will be added on the right hand side.  s

```{r}
cort_clean_demo <- left_join(cort_data_clean, demo_data, by = "Participant_id")
```

Remember, our demo data has 100 subjects, but our clean cort data only has 30 subjects. 
By using **left_join()**, only the data on the left hand size (30 subjects) will be carried over into the new dataset. Let's check:
```{r}
nrow(cort_clean_demo)
```


If we wanted to keep all subjects, we can switch the order of data, so that demo_data is our 'base', and we add cort data to the right side. 
```{r}
cort_all_demo <-  left_join(demo_data, cort_data_clean,
                            by = "Participant_id")

nrow(cort_all_demo)

```

We won't use this data though, because it just has many rows with missing data.
```{r}
tail(cort_all_demo)

```

## 4. Reshape data from wide to long

Also called 'pivoting' data, when you convert a dataset from wide format to long format.

The data is currently in "wide" format, with each row representing a participant. 
However, we have  repeated observations of cortisol (time of day: morning, evening) and days (day 1, day 2). 

Some statistical analyses require long format, where each row is an individual observation, and participants have as many rows as they have observations.  

**pivot_longer()** is a tidyverse function that transforms several columns into a single column, creating a long dataset. 
```{r}
cort_long <- cort_clean_demo %>%
  # columns that have repeated data --> will become a single row
  pivot_longer(cols = starts_with("log"), 
               # name of the new label column (waking, evening)
               names_to = "time_of_day", 
               # prefix to remove from the labels 
               names_prefix = "log_cort_", 
                # name of the new value column (which will contain the cort values)
               values_to = "log_cort")
               
```

We can see that the first several variables are repeated across several rows.
Note: the new (long) variables have been added to the far right of the dataframe.
```{r}
head(cort_long)
```

There are now 120 rows, because each subject (30)  has 4 cort observations.
```{r}
nrow(cort_long)
```


### Side note: string manipulation

We still haven't made a variable for day (day 1, day2).
We need to  do some string manipulation to  separate the 'time of day' from the day number. 

**parse_number()** extracts only the numeric values from a string.

**str_extract()** extracts characters from a string, depending on your specification.
```{r}
cort_long <- cort_long %>%
  mutate(day = parse_number(time_of_day), 
         # [a-z]+ means that we will keep any letters from a-z
         time_of_day = str_extract(time_of_day, "[a-z]+"))

head(cort_long)

```


## 5. Reshape the data from long to wide 

Let's restore the original 'wide' format of the data using **pivot_wider()**
```{r}

cort_wide <- cort_long %>%
  # where to obtain the names / labels for  the new wide columns
  pivot_wider(names_from = c(time_of_day, day), 
               # prefix to be  added to the new wide columns
                names_prefix = "log_cort_", 
              # values of the new wide columns
              values_from = c(log_cort))

nrow(cort_wide)
```

```{r}
head(cort_wide)
```

To check that the wide data has been correctly restored, 
is the new cort variable identical to the original cort variable?

```{r}
#identical(new, original)
identical(cort_wide$log_cort_waking_1, cort_data_clean$log_cort_waking1)
```


Save your cleaned data!
```{r}
# this is the cort in long format, with demo added. 
write.csv(cort_long, "../clean_data/cort_clean_demo_long.csv", row.names = F)
```


