---
title: "Analyzing data"
subtitle: "Lesson 7: Modeling data"
date: "September 2020"
output: 
  html_document:
    theme: flatly
    highlight: haddock 
    df_print: kable
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: true
---

## Goals of this lesson: 

1. Learn about functions and syntax for models in R 
2. Practice examples of different model types 
3. Understand how to obtain model results in R  
4. Learn how to visualize the results of the models

## References
This lesson was adapted from the Columbia University Psychology Scientific Computing [Introductory R Tutorial](https://cu-psych-computing.github.io/cu-psych-comp-tutorial/tutorials/r-core/). 


## 0. Load packages 

We will install a few new packages:
- `effects` for obtaining the fitted effects of model results easily.
- `car` for more useful statistics functions, like Anova()
- `broom` for getting model output into tidy tables.

```{r, message = F}
#install.packages("effects")
# install.packages("car)
# install.packages("broom")

library(tidyverse)
library(effects)
library(car)
library(broom)

```


## 1. Load Data

Load your cleaned data with cortisol and demographics.
```{r, message = F}
# this is the cort in long format, with demo added. 
demo_df <- read_csv("../data/demo_data.csv")
mri_df <- read_csv("../data/mri_data.csv")
```


Merge these datasets by participant_id, and convert intervention to a factor 

```{r}

merge_df <- mri_df %>%
   left_join(., demo_df, by = "Participant_id") %>%
   mutate(intervention = as.factor(intervention),
          age.c = age - mean(age, na.rm = T))


glimpse(merge_df)
```

## 2. Basics of Modeling in R 

### Basic models

For each model type, there is a different function. Many of these are available in base R.

T-TEST: `t.test()`
CORRELATION: `cor.test()`
LINEAR REGRESSION: `lm()`

Each function has their own set of arguments, based on the options available. 

The first argument is always the model formula, which follows a standard syntax.

### Basic model formula

For any type of model with an IV and DV, there is a common syntax. The "~" delineates the direction of the model, where the DV is always on the left of the ~ and the predictors are always on the right.

`Y ~ X`

Example of *t*-test: Does brain volume1 differ by intervention group?

```{r}
t.test(brain_volume1 ~ intervention, data = merge_df)
```

Example of linear regression: Does age predict brain_volumes1?

side note: this output isn't that useful, more on that below.
```{r}
lm(brain_volume1 ~ age, data = merge_df)

```


### To add multiple variables
(in ANOVA or linear regression), simply add  a "+" for each variable on the right of the formula:

`Y ~ X1 + X2 + X3`

Example of linear model example with 2 predictors (continuous and categorical): Does intervention group predict brain_volume1, controlling for age?

```{r}
lm(brain_volume1 ~ age + intervention, data = merge_df)

```

### Interactions

When you want to test the interaction between two variables, add a ":". 
The colon will automatically create an interaction regressor between X1 and X2.
You still need to add each variable alone to calculate their main effects. 

`Y ~ X1 + X2 + X1:X2`

Alternatively, a shortcut is to use an asterisk "*" that provides a shortcut. It will automatically create an interaction regressor AND the main effects for each variable.

`Y ~ X1*X2` is identical to `Y ~ X1 + X2 + X1:X2`

Example of an interaction between 2 continous variables in linear regression: Does age interact with intervention group to predict brain_volume1?

Side note: recommend using mean-centered variables for interactions! 
```{r}
lm(brain_volume1 ~ age.c*intervention, data = merge_df)
```

In the case where there is no IV/DV (i.e. correlation between 2 continuous variables), there is no "~".

Instead there is a ","" to delineate 2 different variables.

Example of correlation: Does brain volume1 correlate with brain volume 2? 

```{r}
cor.test(merge_df$brain_volume1, merge_df$brain_volume2)
```

## 4.  Output of models 

The output of your code will depend on the model type. Linear models have much more information to output relative to t-tests, so they require an additional "summary()" call on the model to see the full results. 

### Linear models

The output includes the following:

a) the formula of your model 
b) residuals
c) coefficients of the model (estimate, st. error, t-value, p-value)
d) overall model results (RSE, R-squared, F-statistics, p-value)

```{r}
# first, we will save our model object and call it lm1
lm1 <-lm(brain_volume1 ~ age + intervention, data = merge_df)

# then we will ask for the summary of the model results
summary(lm1)

# save the model 
save (lm1, file = "my_model.Rda")
```

We can index specific components of the output. What if I want to save the coefficients to a table?

```{r}
summary(lm1)$coefficients

#or, even better: automatically get a tidy table from the broom package
tidy(lm1)

# save the table for later
my_table <- tidy(lm1)

save(my_table, file = "my_table.Rda")
```


## 5. Visualizing models 

There are many ways (and packages) to plot the output of your models. 

We will start with a handy package, [`effects`](https://cran.r-project.org/web/packages/effects/effects.pdf){target="_blank"},  that extracts the 'effects' for you from linear models.

### Visualizing effects of categorical variables from lm

We'll get the effect of intervention  on brain_volume1 from our linear model. This represents the effect of intervention group, controlling for age.

```{r}
# This function requires the X variable in quotes, and the saved model object
intervention_effect <- data.frame(effect("intervention", lm1))

# returns a dataframe with 2 rows, with the fitted effects for each group in the intervention variable.
# also includes standard error, and 95% Confidence Intervals
intervention_effect
```

Let's make plot the effect of intervention group on brain_volume1, with bar graph and error bars. 

```{r}
interventionPlot <- ggplot(data = intervention_effect, aes(x = intervention,  y = fit)) + 
   geom_bar(stat = "identity", aes(fill = intervention)) + 
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) + 
   labs (y = "Brain Volume 1", x = "Intervention") +  
   theme_bw()

interventionPlot
```

Add raw data points. There aren't that many points, so we'll use `geom_point()`  instead of `geom_jitter()`.

We also don't need the legend, so we'll remove it.
```{r}

interventionPlot + geom_point(data = merge_df, aes(x = intervention, y = brain_volume1), alpha = 0.3) +
   theme(legend.position = "none") 
```

### Visualizing effects of continuous variables  from lm 

We'll get the effect of age on brain_volume1 from our linear model. This represents the effect of age, controlling for intervention group. 

```{r}
# This function requires the X variable in quotes, and the saved model object
age_effect <- data.frame(effect("age", lm1))

# returns a dataframe with age at 5 levels (min, -1SD, mean, +1SD, max) 
# with fitted "brain_volume1", standard error, and 95% Confidence Intervals
age_effect
```

Let's make plot the linear effect of age on brain_volume1

`geom_ribbon` is just like `geom_errorbar`, but for linear effects instead of bars.

It allows us to add a shaded region with the 95% CI, indicated by upper & lower bounds.

```{r}
ageplot <- ggplot(data = age_effect, aes(x = age,  y = fit)) + 
   geom_line(size = 1, color = "blue") + 
   geom_ribbon(aes(ymin = lower, ymax = upper), fill = "blue", alpha = 0.2) + 
   labs (y = "Brain Volume 1", x = "Age (years)", title = "Effect of age on brain volume 1") +  
   theme_bw()

ageplot
```

Let's add the raw data. We see how the y axis automatically re-scales to accomodate the data points.
```{r}
ageplot + geom_point(data = merge_df, aes(x = age, y = brain_volume1), alpha = 0.5)

```

## Side notes:

### T-test in R
Default of t-test in R is to perform a Welch's t-test that assumes variances are *not* equal between groups.

To change the assumptions to match SPSS defaults, add the arugment `var.equal = TRUE`
```{r}
t.test(brain_volume1 ~ intervention, data = merge_df, var.equal = TRUE)

```

### Anova in R
There are many packages to do anova in R (too many) all with different assumptions, and they are difficult to keep straight. 

To simplify your life using R, consider your anova in terms of a linear model. 

Remember our linear model with 1 categorical and 1 linear predictor? 
```{r}
summary(lm1)
```

This could be considered an ANCOVA, or an ANOVA with a continuous covariate.

With the `car` package in R, we can actually just take our linear model results and get your results in an ANOVA framework.

type = "III" refers to using type 3 sums of squares, which is generally the default in SPSS. 

*make sure to use `car::Anova()` and not other anova functions.
```{r}
car::Anova(lm1, type =  "III")
```
