---
title: "Lesson 6: Plotting with ggplot"
subtitle: "Materials from CU Psych Scientific Computing Workshop"
date: "September 2020"
output: 
  html_document:
    theme: flatly
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: true
---

## Goals of this lesson:
- recode variables with  `recode()`
- Basic ggplot syntax 
- Example plot types
- Plot Formatting
- Saving plots

## What's `ggplot2`?

ggplot2 is a package in R that allows for highly customizable and pretty plots! Here, were going to learn a few basics of making plots using `ggplot2` that will hopefully get you well on your way to making informative and beautiful data visualizations!

## Links to Files
This lesson was adapted from the Columbia University Psychology Scientific Computing Workshop. The files for all tutorials can be downloaded from [the Columbia Psychology Scientific Computing GitHub page](https://github.com/cu-psych-computing/cu-psych-comp-tutorial){target="_blank"} using [these instructions](/accessing-files/){target="_blank"}. The original file is located here: `/content/tutorials/r-core/4-plotting/index.rmd`.

# 0. Load Packages 

load `tidyverse`, which includes `ggplot2`

```{r, include=FALSE}
library(tidyverse)
```

## 1. Load in sample data

We're going to practice here on a dataset from the 1990 NHANES (National Health and Nutrition Examination Survey). The variables are below. 

Region  -  Geographic region in the USA: Northeast (1), Midwest (2), South (3), and West (4)

Sex     -  Biological sex: Male (1), Female (2)

Age     -  Age measured in months (we'll convert this to years below)

Urban   -  Residential population density: Metropolital Area (1), Other (2)

Weight  -  Weight in pounds

Height  -  Height in inches

BMI     -  BMI, measured in kg/(m^2)

```{r}
nhanes <- read_csv("../data/NHANES1990.csv")
head(nhanes)
```

### Cleaning up the data
Let's `recode()` variables, so that numeric codes are changed to character labels

Side note: we need to add `dplyr::` infront of `recode()` because many different packages have a function called "recode", and we need to tell R which version to use. `dplyr` is a package in the tidyverse. 
```{r}
nhanes <- nhanes %>%
    mutate(Age = Age/12, # convert age to years
          Urban = dplyr::recode(Urban, '1' = 'Metro Area', '2' = 'Non-Metro Area'),
          Region = dplyr::recode(Region, '1' = 'Northeast', '2' = 'Midwest', '3' = 'South', '4' = 'West'), 
          Sex = dplyr::recode(Sex, '1' = "Male", '2'="Female"))

head(nhanes)
```

## 2. Basic `ggplot2` Syntax

-`ggplot()` command usage including saving to a variable with `a <- ggplot()`

-The general format is `ggplot(data, aes(x = [x variable], y = [y variable])` 
    - x and y variables are always specified in this `aes()` subfunction

Run this line of code:

```{r}
ggplot(nhanes, aes(x = Age, y = Weight))
```
  
  * Axes are set up the way we'd expect, and seem to have sensible values
  * Why is nothing on this graph yet? we haven't put any graphic actually on the axes yet

`We need to tell the ggplot() call what kind of graphic to put on the axis`

  * A lot of the time, the syntax is `geom_[something]`

## 3. Example Plot Types

### Scatter plots

Let's do scatter with `geom_point()` first.

```{r}
ggplot(nhanes, aes(x = Age, y = Weight)) + 
  geom_point()
```

Wow, lots of data points! 

We can make the points smaller with `size` and make them slightly transparent with `alpha`.

```{r}
ageWeightPlot <- ggplot(nhanes, aes(x = Age, y = Weight)) + 
  geom_point(size = .1, alpha = .3)

ageWeightPlot
```

Next, we can change the color of the data points by Sex. 

We put `color = Sex` into the `aes()` function because it's a grouping factor.


```{r}
ageWeightPlot <- ggplot(nhanes, aes(x = Age, y = Weight)) + 
  geom_point(size = 1, alpha = .3, aes(color = Sex))

ageWeightPlot
```

### Histograms

Test out changing bins # and see how the histogram changes.

Side note: For spaces that can be filled in with color (e.g. histogram), we use the term `fill` instead of `color`.

```{r}
weightHistogram <- ggplot(nhanes, aes(x= Weight)) +
  geom_histogram(bins = 100, fill = 'dark blue')
weightHistogram
```

Do you hate the gray background? We can change the plot aesthetics with `themes()`

More themes can be found here: https://ggplot2.tidyverse.org/reference/ggtheme.html

```{r}
weightHistogram <- ggplot(nhanes, aes(x= Weight)) +
  geom_histogram(bins = 100, fill = 'dark blue') + 
  theme_bw()
weightHistogram
```

Let's fill the histogram colors by Sex. 

`scale_fill_manual()` allows us to add customized colors instead of the default colors when using more than one color.

More on color palettes can be found here: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/
```{r}
weightSexHistogram <- ggplot(nhanes, aes(x= Weight)) +
  geom_histogram(bins = 100, aes(fill = Sex), alpha = 0.5) + 
  scale_fill_manual(values=c("dark red", "dark blue")) + 
  theme_bw()
weightSexHistogram
```

### Boxplots 

Let's compare the distributions of weight, by sex, side by side with a boxplot.

In ggplot, the boxplot automatically calculates the quartiles of the distribution (not mean and sd). 
So the center line represents the median, and the box includes the 2st and 3th quartiles.

```{r}
weightbySexBox<- ggplot(nhanes, aes(x= Sex, y = Weight)) +
  geom_boxplot(aes(fill = Sex), alpha = 0.5)  +
  scale_fill_manual(values=c("dark red", "dark blue")) + 
  theme_bw()

weightbySexBox

```

### Summary plots

What if we want to plot the mean and standard deviation for Males and Females?

First, let's make a summary table 
```{r}
weight_table <- nhanes %>%
  group_by(Sex) %>%
  summarize(mean_weight = mean(Weight), 
            sd_weight = sd(Weight), 
            upper = mean_weight + sd_weight, 
            lower = mean_weight - sd_weight)

weight_table
```

#### bar-graph summary plot

`geom_bar()` will make a bar plot.

`stat = "identity"` means that we want to use the exact values we provided in the table. 

`geom_errorbar()` takes two main arguments: ymin and ymax, and width sets the appearance of the bars 
```{r}
# notice that we changed the data - using summary table instead of raw data.
weightbySexBar <- ggplot(data = weight_table, aes(x = Sex, y = mean_weight)) +
   geom_bar(aes(fill = Sex), stat = "identity",alpha = 0.5) + 
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.5) + 
   scale_fill_manual(values = c("dark red", "dark blue")) + 
  theme_bw()

weightbySexBar

```


#### dot-plot summary plot

new argument: `ylim()` will set the y axis range (it's good practice to start at zero)

Side note: here, we use 'color' instead of 'fill' because they are dots and lines, not spaces to fill. 
```{r}

weightbySexDot <- ggplot(data = weight_table, aes(x = Sex, y = mean_weight)) +
               geom_point(aes(color = Sex), size = 2) + 
               geom_errorbar(aes(color = Sex, ymin = lower, ymax = upper), width = 0) + 
               scale_color_manual(values = c("dark red", "dark blue")) + 
               ylim(0, 300) + theme_bw()

weightbySexDot
```

Let's add the raw data points with `geom_jitter()`. 

Test out what happens when you change the width and weight aguments.
```{r}

weightbySexDotRaw <- ggplot(data = weight_table, aes(x = Sex, y = mean_weight)) +
  # add jittered points of raw data with lots of transparency, because there are so many points.
   geom_jitter(data = nhanes, aes(x = Sex, y = Weight),
              alpha = 0.1, size = .1,height = 0, width = 0.1, color = "gray") + 
  # this code is identical to the plot above, we are just layering it on.
               geom_point(aes(color = Sex), size = 2) + 
               geom_errorbar(aes(color = Sex, ymin = lower, ymax = upper), width = 0) + 
               scale_color_manual(values = c("dark red", "dark blue")) + 
               ylim(0, 300)  + theme_bw()

weightbySexDotRaw
```

Side note: "Removed 121 rows containing missing values" indicates that we are cutting-off our data. 
Now that we've added raw data, the range of values for Weight is much wider. Try expanding ylim.


## 4. Plot Formatting

### Titles and axis labels

The `labs()` command can be added to `ggplot2` with different arguments, lke `x`, `y`, or `title` to make the plots clearer.

```{r}
# we'll leave X blank, becasue we already have useful tick labels.

weightbySexDotRaw + labs(x = ' ', y = 'Weight in Pounds', title = 'Sex differences in weight')
```


### Facetting plots

Does the relationship between Age and weight differ by sex? 

We can use the `facet_wrap()` function to look at Males and Females separately.

```{r}
facetPlot <- ggplot(nhanes, aes(x = Height, y = Weight)) + 
  geom_point(alpha = .1, size = .1) + 
  facet_wrap(~Sex) + theme_bw()

facetPlot
```


## Saving to Files

`ggsave()` includes arguments for `file` , `plot`, `dpi` (resolution), `width`, and `height`.

We can use a variety of file formats (png, pdf, jpeg, tiff).

```{r, eval = FALSE}
ggsave('newPlotTest.pdf', plot = facetPlot, dpi = 300, width = 5, height = 5)
```

