---
title: "Analyzing data: modeling"
subtitle: "Bonus lesson: mixed effects models with longitudinal data"
date: "Sept, 2020"
output:
  html_document:
    number_sections: no
    df_print: kable
    code_folding: show
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

## 0. Load packages 

```{r, message = F}
# install.packages("lme4")
# install.packages("lmerTest")

library(tidyverse)
library(lme4)
library(lmerTest)
```


## 1. Load Data

load the cort_clean_demo_long.csv data
```{r}

cort_long <- read_csv("../clean_Data/cort_clean_demo_long.csv")
```

prep data: before modeling, it's a good idea to change your character variables into "factors" so that R knows they are categorical / factor variables.
```{r}
cort_long <- cort_long %>%
  mutate(intervention = as.factor(intervention), 
         time_of_day = as.factor(time_of_day), 
         sex = as.factor(sex))

head(cort_long)

```

## 2. visualize the data

we actually have within-subject observations in this dataset: time of day (waking, evening) and day (1,2).

Let's plot the cortisol slope across the day, and facet by day of cort collection.

`geom_line()` allows us to add a "grouping" variable, so we can connect the lines by Participant_id.

We can still color by intervention.
```{r}

ggplot(data = cort_long, aes(x =time_of_day, y = log_cort)) + 
  geom_point(aes(color = intervention)) + geom_smooth(method = "lm") +
  facet_grid(~day) + theme_bw() +
  geom_line(aes(group = Participant_id, color = intervention)) + 
  labs(x = "", y = "Cortisol Value (log)")

```

side note: ggplot automatically orders factors alphabetically (bedtime, waking).

Let's swap the order of the time_of_day factor using `relevel()` so that waking appears first.

```{r}
cort_long <- cort_long %>%
  # the relevel function changes the "reference" level of the factor
  mutate(time_of_day = relevel(time_of_day, "waking"))

ggplot(data = cort_long, aes(x =time_of_day, y = log_cort)) + 
  geom_point(aes(color = intervention)) + geom_smooth(method = "lm") +
  facet_grid(~day) + theme_bw() +
  geom_line(aes(group = Participant_id, color = intervention)) + 
  labs(x = "", y = "Cortisol Value (log)")


```


## 3.  Model the longitudinal data 

Let's create a model of this data, where we can look at how cortisol slopes vary by intervention group.

Let's use **mixed effects modeling**  with random effects to account for within-subject variance, using the `lme4` and `lmerTest` packages.

We will have two random effects: within-subject slopes (time of day) and per-subject intercepts (waking level)

```{r}

lmer_mod1 <- lmer(log_cort ~ intervention +  time_of_day + # variables of interest
                    day +age + sex + # covariates 
       (1 + time_of_day | Participant_id),  # random effects
       data = cort_long) # data

summary(lmer_mod1)
```


That output is a bit much, so let's pull out just the coefficients of the fixed effects. 
```{r}
fixed_effects <- summary(lmer_mod1)$coefficients
fixed_effects
```


We can also get the individual intercepts and slopes from each subject (random effects)
```{r} 
random_effects <- ranef(lmer_mod1)
head(random_effects)
```

