---
title: "Merging and reshape data: challenge"
date: "Sept , 2020"
output:
  html_document:
    number_sections: no
    df_print: kable
    code_folding: show
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

## key concepts:
- Merge datasets together with **join()**
- Reshape data from wide to long with **pivot_longer()** 
- Reshape data from long to wide withs **pivot_wider()**

## 0. load packages 
```{r, warning = F, message = F}
# load packages 
library(readxl)
library(tidyverse)
```

## 1. load data 
open the demographic data and a new dataset, mri_data.csv
```{r}

```


## 2. Merge the  data

merge the demo and mri data using 'Participant_id' as the key. 

use left_join(), with the demographic data as the base, 
and mri added to the right side

name the new dataframe 'mri_demo'
```{r}
```

## 3. Reshape the data from wide to long

Repeated variables: brain_volume1, brain_volume2, brain_volume3 

create a long dataset called "mri_long"
with a column named "timepoint" with values 1,2,3
and a column named "brain_volume" that contains the values of the brain volume variables.
```{r}

mri_long <- mri_demo  %>%
  pivot_longer(cols = 
               names_to = 
               names_prefix =
               values_to =   )

```

## 4. Reshape the data from long to wide 

Reproduce the wide dataset from the long data you just created.

```{r}
mri_wide <- mri_long %>%
  pivot_wider(names_from =
                names_prefix =
              values_from =    )

```
