---
title: "Analyzing data: plotting challenge"
date: "Sept, 2020"
output:
  html_document:
    number_sections: no
    df_print: kable
    code_folding: show
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---


# 0. load packages 

```{r, warning = F, message = F}
# load packages 
library(readxl)
library(tidyverse)
```


# 1. load data 

load the cleaned cortisol data with demographic data added: cort_clean_demo_long.csv
```{r}

cort_df <- read_csv("../clean_data/cort_clean_demo_long.csv")
head(cort_df)
```

# 2. Make a bar graph with mean cortisol values, by time of day 

First, make the summary table
```{r}
cort_table <- cort_df %>%
  group_by(time_of_day) %>%
  summarize(mean_cort = mean(log_cort), 
            sd_cort = sd(log_cort), 
            upper = mean_cort + sd_cort, 
            lower = mean_cort - sd_cort)

cort_table
```

Next, plot the bar graph with error bars. 

Fill the bars (colors) based on time of day.

Change y and x axis labels and add a title for your plot
```{r}
CortBar <- ggplot(data = cort_table, aes(x =  time_of_day, y = mean_cort)) +
   geom_bar(aes(fill = time_of_day), stat = "identity",alpha = 0.5) + 
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.5) + 
   scale_fill_manual(values = c("dark red", "dark blue")) + 
  theme_bw() + 
  labs(x = "", y = "Log Cort (mean)", title = "Time of Day effect on Cortisol")

CortBar

```

Add raw data points using geom_jitter() 

Hint:  make sure to add the raw data as the first argument!

```{r}
CortBarJitter <- CortBar + geom_jitter(data = cort_df, aes(x = time_of_day, y = log_cort, fill = time_of_day), 
                      width = 0.1, height = 0, alpha = 0.5)

CortBarJitter
```


## 3. Facet the time-of-day plot by Intervention group.

First, make  a new summary table, where we also group by intervention.
```{r}
cort_table <- cort_df %>%
  group_by(time_of_day, intervention) %>%
  summarize(mean_cort = mean(log_cort), 
            sd_cort = sd(log_cort), 
            upper = mean_cort + sd_cort, 
            lower = mean_cort - sd_cort)
cort_table
```

Next, make the same bar plot, but this time add an extra line with facet_wrap()

```{r}

CortBar <- ggplot(data = cort_table, aes(x =  time_of_day, y = mean_cort)) +
   geom_bar(aes(fill = time_of_day), stat = "identity",alpha = 0.5) + 
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.5) + 
   scale_fill_manual(values = c("dark red", "dark blue")) + 
  theme_bw() +
  labs(x = "", y = "Log Cort (mean)", title = "Cortisol by Time of Day and Intervention Group") +
   facet_wrap(~intervention) 

CortBar

```