---
title: "Analyzing data: plotting challenge"
date: "Sept, 2020"
output:
  html_document:
    number_sections: no
    df_print: kable
    code_folding: show
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

# 0. load packages 

```{r, warning = F, message = F}
# load packages 
library(readxl)
library(tidyverse)
```


# 1. load data 

load the cleaned cortisol data with demographic data added: cort_clean_demo_long.csv
```{r}

```

# 2. Make a bar graph with mean cortisol values, by time of day 

First, make the summary table called 'cort_table'
```{r}
cort_table <- 
```

Next, fill in the code below to plot the bar graph with error bars. 

Fill the bars (colors) based on time of day, and add your own custom colors.

Change y and x axis labels and add a title for your plot
```{r}
CortBar <- ggplot(data = cort_table, aes(    )) +
   geom_bar(aes(      ),      ) + 
   geom_errorbar(aes(        ),      ) + 
   scale_fill_manual(values = c("    ", "     ")) + 
   theme_bw() + 
   labs(     )

CortBar

```

Add raw data points using geom_jitter() 

Hint:  make sure to add the raw data as the first argument!

```{r}
CortBarJitter <- CortBar + geom_jitter(data =       , aes(       ),       )

CortBarJitter
```


## 3. Facet the time-of-day plot by Intervention group.

First, make  a new summary table, where we group by time of day and intervention.
```{r}
cort_table <- 
```

Next, make the same bar plot, but this time add an extra line with facet_wrap()

```{r}

CortBarFacet <- 

  

CortBarFacet

```