---
title: "Manipulating data in R: challenge"
date: "Sept , 2020"
output:
  html_document:
    number_sections: no
    df_print: kable
    code_folding: show
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

## key concepts:
- View data with **hist()**
- Exclude data with **filter()**
- Create new variables with **mutate()**
- Change variable names with **rename()**
- Subset variables with **select()**


## 0. Open packages 
```{r, warning = F, message = F}
# load packages 
library(readxl)
library(tidyverse)
```

## 1. load the data

Open the  demographics data

```{r}
demo_data <- read_csv("../data/demo_data.csv")
```

## 2. Inspect the data
visualize age with a histogram
```{r}
hist(demo_data$age)
```

## 3. clean / filter the data

filter the data to only include males over the age of 30. 
save the cleaned data as "demo_clean"
```{r}
demo_clean <- demo_data %>%
  filter(sex == "Male" & age > 30)

```

how many participants are in your cleaned dataset? 
```{r}
nrow(demo_clean)
```

## 4. Create new variables 

Create a new variable called age_squared, which is the square of age.
hint, the expression in R for squaring is "^2"
```{r}
demo_clean <- demo_clean %>%
  mutate(age_squared = age^2)

```

make a histogram of the new variable
```{r}
hist(demo_clean$age_squared)

```

##  5. Rename variables

Rename the variable "intervention" to "Group"
```{r}
demo_clean <- demo_clean %>%
  rename(Group = intervention)

```

## 6. Select variables

create  a dataframe that only has 3 columns, Participant_id, age, and age_squared.
save this dataframe as "age_data" 
```{r}
age_data <- demo_clean %>%
  select(Participant_id, age, age_squared)
  

```