---
title: "Summarizing data in R: challenge"
author: "Michelle VanTieghem"
date: "Sept , 2020"
output:
  html_document:
    number_sections: no
    df_print: kable
    code_folding: show
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

For this challenge, you're going to learn how to summarize a new dataset.
We will use an example cort dataset generated for this course
filepath: "../data/cort_data_wide.csv"

# 0. Open packages 
```{r, warning = F, message = F}
# load packages 
library(readxl)
library(tidyverse)
library(lubridate)
```

# 1. load the data
add the file path to the read_csv() function
```{r, message = F, warning = F}
cort_data <- read_csv("../data/cort_data_wide.csv")
```


# 2. Inspect the data 

How many rows and columns does this dataset have?
hint: use nrow() and ncol()
```{r}
nrow(cort_data)
ncol(cort_data)
```

What are the column names of the data?
hint: use  names()
```{r}
names(cort_data)

```

Take a glimpse of the data
```{r}
glimpse(cort_data)

```

There are a lot of variables in this  dataset! 
For now, we will focus on the cortisol variable 'cort_waking1', 
and the grouping variable 'caregiver_type'.


#  3. Summarizing data 

### Are there missing cort values?
hint: use sum(is.na())
```{r}
sum(is.na(cort_data$cort_waking1))
```

### using summarize(), calculate the mean and SD of variable called 'cort_waking1'
hint: remember to remove NA values with na.rm  = T
```{r}
cort_data %>%
  summarize(mean = mean(cort_waking1, na.rm = T), 
            sd = sd(cort_waking1, na.rm = T))
```

### using group_by(), calculate how many participants had which 'caregiver type'
hint: you have 2 options, use count() or summarize(n = n())
```{r}
cort_data %>%
  group_by(caregiver_type) %>% 
  count()

```

### combine these tables: 
use group_by() and summarize() to compute the mean and sd cortisol values by caregiver type.
hint: don't forget to remove NAs with na.rm = T
```{r}
cort_data %>%
  group_by(caregiver_type) %>%
  summarize(n = n(), 
            mean =  mean(cort_waking1, na.rm = T), 
            sd = sd(cort_waking1, na.rm = T))

```

### save this table to csv format using write.csv()
hint: don't forget the argument row.names = F
```{r}


```